Entropy, Symbolization and Mutual Information \\ Teorical-Numerical Quarentine Dissertation

### 2020-06-18 Reportes/Reporte20200601.tex
reporte extendido

We have established that the method of counting the discrete values of
the measured LFP is the one that gives a trustworthy estimation of the
probabiliy function. In some sense, it is a good representative
of the probabilites of the \emph{measured values} of the data.
As a side effect, it also helps us to reveal numerical artifacts
on the data. Without a relieable measure of the Probabilites
we cannot  calculate causality functions, such as Mutual Information and
Transfer Entropy, which are our true aim.
Could we use the countmap method for such calculations?
The answer is no. The problem is that the probabilites of joint events
decrease as a power of the number of events that we are measuring: the data
points get diluted in a higher dimensional space. We are seeing
discrete events that are sampled from a process that has around 1000 symbols.
They appear to be more or less normaly distributed, with a standard deviation
of around 12 simbols. So around the central 50 simbols we have 95 percent of all
events (the 68-95-99 rule of gaussian distributions).
In a very rough Fermi approximation these symbols could be thought of being
equiprobable. If we measure the joint probability of two channels,
the appareance of a certain pair
of symbols in those ranges is around 1/2500. We have around 2000 samples in
the example experiment. That means that we are going to observe each pair
at most once, which makes a very poor statistic indeed.
If that were not bad enough,
the central symbols account mostly for the background and thermal noise.
Those are the more probable pairs of symbols that we are going to see: the
ones corresponding to spikes are very rare, and to see
two correlated spikes is going to be an extremely rare event with this
resulution and number of samples.

@@

@@

Another caveat, often overlooked, is that Entropy of an array of symbols is invariant to ordering. One could in principle regard a signal as a sampling of random variables, reorder them, and obtain the same distribution, and thus, the same entropy. As an example, we could take every letter, punctuation mark and space of this paragraph, garble them into nonsense, and claim that they have the same Entropy as the original paragraph, and therefore, the same Information, which is as much nonsense as the random letters. But actually this is done implicitly in a lot of numerical estimations of Entropy.

1
\bibliography{./BiblioReportes01}{}
\bibliographystyle{plain}

>>>>

We have established that the method of counting the discrete values of
the measured LFP is the one that gives a trustworthy estimation of the
probabiliy function. In some sense, it is a good representative
of the probabilites of the \emph{measured values} of the data.
As a side effect, it also helps us to reveal numerical artifacts
on the data. Without a relieable measure of the Probabilites
we cannot  calculate causality functions, such as Mutual Information and
Transfer Entropy, which are our true aim.
Let us recall the definition of the first measure. Let us say that
$y$ is a result of the random process $Y$ and $x$ of the random
process $X$ (for simplicity in the notation we shall identify
the process with its encoding alphabet, and from now on we shall
restrict ourselves to discrete process). Mutual Information
between process $Y$ and $X$ is given by \cite{Cover1991}:
\begin{equation}
  I(X;Y)=\sum_{x \in X} \sum_{y \in Y} P(x,y) \log \frac{P(x,y)}{P(x)P(y)}
\end{equation}
where $P(y,x)$ is the probability of the joint event and the other are
individual probabilities. Here is where our problem grows in dimensionality.
These joint probabilities have as domain a cartesian product of the
separated spaces $X$ and $Y$. If one has $n$ symbols in one, and
$m$ symbols in the other, the joint space alphabet would have $ n \times m$ 
symbols.
Could we use the countmap method for such calculations?
The answer is no. The problem is that the probabilites of joint events
decrease as a power of the number of events that we are measuring: the data
points get diluted in a higher dimensional space. We are seeing
discrete events that are sampled from a process that has around 1000 symbols.
They appear to be more or less normaly distributed, with a standard deviation
of around 12 simbols. So around the central 50 simbols we have 95 percent of all
events (the 68-95-99 rule of gaussian distributions).
In a very rough Fermi approximation these symbols could be thought of being
equiprobable. If we measure the joint probability of two channels,
the appareance of a certain pair
of symbols in those ranges is around 1/2500. We have around 2000 samples in
the example experiment. That means that we are going to observe each pair
at most once, which makes a very poor statistic indeed.
If that were not bad enough,
the central symbols account mostly for the background and thermal noise.
Those are the more probable pairs of symbols that we are going to see: the
ones corresponding to spikes are very rare, and to see
two correlated spikes is going to be an extremely rare event with this
resulution and number of samples.

@@

\section{Symbolization strategies}

The problem with the histograms is that we induce a certain partition of the
data, and there it is a certain degree of arbitrairness. We have also seen
that partitions that are too fine would be useless for the higher dimensional
measures. We shall test some symbolization strategies that do not depend
on the exact values of the data, as we have also tested that some of these
values may be unreliable due to measurement artifacts.
Steuer \emph{et al.} \cite{Steuer2004} used a two symbol system based on
the finite time derivative of the signal. I shall use the same basic idea
but with three symbols, using $1$ for increasing signal, $0$ for stationary, and
$-1$ for decreasing signal.

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=1.0\textwidth]{entropia2s01.png}
  \end{subfigure}
   \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=1.0\textwidth]{histo2s01.png}
  \end{subfigure}
   \caption{The entropy for 3 symbols over all channels  and its histogram. The later was done using only the putative active channels. No discernible features are visible,
     except the unusable channels, that have zero entropy. Compare with figure
     \ref{entrocomp}. }
   \label{entro2s01}
\end{figure}

In the figure \ref{entro2s01} we show the results of this symbolization procedure.
As one can see, there is no range and therefore no contrast between the
active channels and the inactive ones. This symbolization is not adecuate. The
noise, which varies rapidly between increasing and decreasing signals, drowns
out all features of neuronal activity when reduced to such a simple characteristic.
Steuer \emph{et al.} do not provide clues of the timestep they used, but the caveat
is there. I propose to extend the alphabet in the following manner.
Instead of considerin one time step, which is a sampling step of the BrainWave
BioCAM, we shall use time windows that are of the order of magnitud of physiological
events, around 1ms. In each window we shall calculate the sums of the signs
of the differences, that is, sum the signs of the former symbolization. The new symbols
will be a measure of how much was moving the signal in the positive or negative
direction, discretized in reasonable amounts. If a window contains mostly noice,
the sum will be near zero. Otherwise it shall depart notably from zero
in a window which encompases the polarization or depolarization of a neuron. A window
may encompass $n$ frames, so the alphabet shall range from $-n$ to $n$, having
$2n+1$ symbols. We shall choose the $n$ value as the ones which provides the
higher contrast between entropies, that is, the largest range of entropies between
active and inactive channels. The largest spread of entropy values ignoring the
unusable channels appears at $n=8$, which at this sampling rate is a little longer
than one milisecond. This could be a good symbolic alphabet for obtaining the
Mutual Information, although it is still quite large with 17 symbols. 
\begin{figure}
  \begin{center}
    \includegraphics[width=0.85\textwidth]{comparasumaentropias01.png}
    \end{center}
  \caption{Comparission of the summed symbol entropies. Each image corresponds
    to summing over the stated number of symbols on the header.}
  \label{compsumentro}
\end{figure}

In the figure  \ref{h8sim} we
show the distribution of symbols for the sum over 8 frames, ranging from
$-16$ to $+16$. The distribution seems consistent across channels, having a
gross normal appearence with richer structure therein.

\begin{figure}
  \centering
  \includegraphics[width=3.0in]{histogramas_8simbolos.png}
  \caption{Histogram for two channels on the putative active Area, coordinates indicated in the legend. Compare with figure \ref{entro2s01}.}
  \label{h8sim}
\end{figure}

@@

Another caveat, often overlooked, is that Entropy of an array of symbols is invariant to ordering. One could in principle regard a signal as a sampling of random variables, reorder them, and obtain the same distribution, and thus, the same entropy. As an example, we could take every letter, punctuation mark and space of this paragraph, garble them into nonsense, and claim that they have the same Entropy as the original paragraph, and therefore, the same Information, which is as much nonsense as the random letters. But actually this is done implicitly in a lot of numerical estimations of Entropy. In the last section
I proposed adding the symbols of a certain time window to obtain a richer symbolization,
and thus describe a larger set of messages. There is more detailed version
of this approach. The idea is to regard finite succesion of symbols as
higher order symbols, in a larger alphabet, a vocabulary in some sense. The entropy
that one obtains calculating $n$ symbol words is called ``n-block Entropy''.
This measure of entropy does not converge as we take larger words. As we have
established, Shanon's Entropy grows as the logarythm of the size of the
alphabet, or in this case, the vocabulary. If the signal is rich in pattens, and
long enough, many words shall appear. What it is done here is to establish
an average entropy increasse per new symbol. If Shannon's Entropy for the $n$ block
vocabulary is denoted $H_n$, the average increase in entropy for each new symbol
will be:
\begin{equation}
    h_n=(H_{n+1}-H_n)    
\end{equation}
This quantity converges under normal circumstances, and its limit is called
the \emph{source entropy}. Unconveniently, it converges very slowly, so in principle
we shall use very large words to obtain a good approximation. This becomes a
problem, because we are constrained by the data set length to obtain adecuate
probabilities in the estimation of this entropy. Let us
think that the data length is $M$ symbols
long, and we are trying to establish $n$ word probabilities. It the alphabet
contians $N$ different symbols, there are $N^n$ possible words. If they were
equiprobable, in order to detect on average once each word, t
he next relationship should hold:
\begin{equation}
  1 \approx M/(N^n)
\end{equation}
that means that to be marginally able to detect an arbitrary $n$ length word,
$n$ shall be at most:
\begin{equation}
  n < \log(M)/\log(N).
\end{equation}
or, equivalently, $ n < \log_N (M) $, that is, the logarythm of the length of
the measured data in the base of the symbol alphabet size. In a concrete case,
if we have $2000$ data points per channel, and I propose a three symbol
alphabet, we cannot realiably measure probabilities beyond $n=7$. As a curious
note, this is very close to the best case in the  ``summed'' symbol that
I proposed on the last section. The similitude may derive in that
the summed symbol is a ``compressed encoding'' of the word symbol, and that
both may cointain similar information, altough on second looks,
the summed symbol, especially if the range  of the sum is relative large, may
encode as the same symbol very different signal segments, and the compression
is not homgeneous. The quantity of words $W_3$ that add up exactly to $m$ using at most
$n$ symbols choosen from the alphabet $S_3={-1,0,1}$ is
\begin{equation}
  W_3=\sum_{j=0}^\nu \frac{n!}{(m-n-2j)!(n+j)!j!}, 
\end{equation}
where $\nu$ is the integer part of $(n-)/2$ and represents the pairs of $+1$ and
$-1$ that add up to zero and don't alter the sum. as an example we can calculate
how many words are commpresed in the 8 symbols sum for different values of $m$.
\begin{center}
  \begin{tabular}{ l | r }
    $m$ & $W_3$ \\
    \hline
    0 & 1107 \\
    1 & 1016 \\
    2 & 784 \\
    3 & 504 \\
    4 & 266 \\
    5 & 112 \\
    6 & 36 \\
    7 & 8 \\
    8 & 1 \\
    \hline
  \end{tabular}
\end{center}
The most extreme case is the $m=0$, where there could be such disimilar symbols
compresed as the sequence $0,0,0,0,0,0,0,0$, a completely flat signal,
and the sequence $-1,-1,-1,-1,+1,+1,+1,+1$, which could be a spike! It is interesting
to note that even so, the entropy thus obtained is concentrated in the
physiological
expected areas, namely CA3. After some pondering, it turns out that
the lossy compression is only effective if we perform the change of dictionary
( the sum over windows) if we do not make a running window, but a time stepped
window. The running window does not lose in general the ternary message, because
after $n$ time steps, we have enough linear equations to recover the
subyacent signs. Therefore, the succesive messages are not independent. They
follow a very specific and strict succesion, and that produces higher entropies,
revealing the structure of the original complex message that is subyacent to the
encoding. Heuristically, that also could point out to the fact that
the optimal summing window has 8 time frames, and the best block entropy
that we can calculate is with 7 time frames.

We should not be to hasty to regard the expanded and detailed vocabulary
as the best simbolization strategy. As before, excess symbols may actually be a
hindrance to a good interpretation of the entropy measure. Let us see the
entropies from 1 to 6 symbols in the figure \ref{entro6symbs}.

\begin{figure}
  \includegraphics[width=0.8\textwidth]{comparaentrobloc01.png}
  \caption{Block Entropies for words with three symbols from the
    set $\{ -1, 0, 1 \}$. 
  }\label{entro6symbs}
\end{figure}

A bit to the surprise and dismay of the author, the Block Entropy for this
data provides less contrast (and therefore less information about the channels)
than the sum of the symbol entropies. A plausble explanation is that the
more detailed symbolic set flattens the distribution, making noisy and
active channels more alike in this encoding. We shall test this numerically
on the next figure, \ref{histo6simbols}. The histogram looks terribly
messy and without any ordering, much in contrast with the one in figure
\ref{h8sim} where a natural ordering seems to be present. Worse yet,
the comparision with a noisey channel doesn't seem to yield any qualitative
differences, but a sort of grouping of the probabilites that seem similar
in both cases, thus indicating that this symbolization is better at
describing the noise than the physiological signal.

\begin{figure}
  \centering
  \begin{subfigure}[b]{\textwidth}
  \includegraphics[width=0.8\textwidth]{histo6simbols.png}  
  \end{subfigure} \\
  \begin{subfigure}[b]{\textwidth}
  \includegraphics[width=0.8\textwidth]{histo6simbruido.png}  
  \end{subfigure} 
  \caption{Distribution for the 6 symbol words in an active
    Channel. The labels are ordered numerically from
    $(-1,-1, -1, -1,-1,-1)$ to $(1,1,1,1,1,1)$. In the above
    plot the distribution for an active channel is shown. Below we see
    a graphic for a noisy channel. 
  }\label{histo6simbols}
\end{figure}

\bibliography{./BiblioReportes01}{}
\bibliographystyle{plain}
